The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-08-29 thesis/rv_models.tex
thesis refining data section
julio edits at durham

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1-3
arcseconds (see REF \citet{tab:rv_measurements}). We combine the mean RV
measurement from each study using the inverse-variance weighted mean
\(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\). We remove stars with significant velocity dispersions
as measured between observations in a study or between studies. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically).

@@

\begin{table*}[t]
\centering
\caption{Summary of velocity measurements and derived properties.}
\begin{tabular}{lllllllll}
\toprule
 & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
\midrule
Scl & combined &  & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
 & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
 & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
 & walker+09 & MMFS & 1818 & 1522 & 1417 & 1328 & 1.8 & 3\\
 & APOGEE & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
UMi & combined &  & 4714 & 1225 & 1148 & 863 & 2.1 & \\
 & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
 & pace+20 & DEIMOS & 1716 & 1538 & 829 & 678 & 2.5 & 1\\
 & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & 2\\
 & APOGEE & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
\bottomrule
\end{tabular}
\end{table*}

>>>>

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1-3
arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
mean RV measurement from each study using the inverse-variance weighted
mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\). We remove stars with significant velocity dispersions
as measured between observations in a study or between studies. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically).

@@

\begin{table*}[t]
\centering
\caption[Spectroscopic LOS velocity measurements]{Summary of velocity measurements and derived properties. }
\label{tbl:rv_measurements}
\begin{tabular}{lllllllll}
\toprule
 & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
\midrule
Scl & combined &  & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
 & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
 & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
 & walker+09 & MMFS & 1818 & 1522 & 1417 & 1328 & 1.8 & 3\\
 & APOGEE & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
UMi & combined &  & 4714 & 1225 & 1148 & 863 & 2.1 & \\
 & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
 & pace+20 & DEIMOS & 1716 & 1538 & 829 & 678 & 2.5 & 1\\
 & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & 2\\
 & APOGEE & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
\bottomrule
\end{tabular}
\end{table*}

measurement

### 2025-08-29 thesis/sculptor.tex
julio edits at durham

\subsubsection{Derived properties and
setup}\label{derived-properties-and-setup}

From the observed properties of Sculptor, we can infer reasonable
\(\Lambda\)CMD dark matter halo hosts. Table~\ref{tbl:scl_derived_props}
reports our inferred halo and kinematic properties of Sculptor. First,
taking the absolute magnitude from \citet{munoz+2018} with the
mass-to-light ratio from \citet{woo+courteau+dekel2008} (1.7 with
\(\sim\) 0.17 dex uncertainty), the total current stellar mass of
Sculptor is \(M_\star \sim 3.1 \times 10^6 \Mo\). Based on the stellar
mass-\(v_{\rm max}\) relation \citep[from][]{fattahi+2018}, see also
\ref{fig:SMHM}, the halo should have
\(v_{\rm max} \sim 31 \pm 3 \,\kms\). Finally, using the
\citet{ludlow+2016} \(z=0\) mass-concentration relation, this constraint
translates into a \(r_{\rm max} \sim 6 \pm 2\,{\rm kpc}\), or
alternatively \(M_{200} = 5\times10^9\) with \(c=13\).

Fig.~\ref{fig:scl_halos} illustrates these estimates visually. The
stellar-mass \(v_{\rm max}\) constraint translates to an estimate of
\(v_{\rm max}\) only (horizontal band). The mass-concentration relation
describes the relationship between \(v_{\rm max}\) and \(r_{\rm max}\),
an approximately diagonal linear band. And finally, we include a curved
line which illustrates where the initial velocity dispersion is
approximantly larger than the present-day observed dispersion
\(v_{\rm circ}(R_h) / \sqrt{3} \sim \sigma_v \gtrsim 9\,\kms\).

The intersection of the constraints on the halo leads us to select an
initial halo, \texttt{compact}, which is more compact than the mean but
with a observationally consistent initial velocity dispersion. We also
have the \texttt{light} halo which occupies the intersection between
3\(\sigma\) lowest mass and the velocity dispersion constraints, which
we come back to in Section~\ref{sec:scl_lmc}.

While there is some range in the choice of initial halo, reasonable
changes to the initial halo do not substantially affect the tidal
evolution for Sculptor. The velocity dispersion, in particular,
translates to an estimate of the mass within \(r_h\). As a result, halos
with the same velocity dispersion only differ in total mass, but the
density slope and circular velocity near the stellar component is
similar. These halos will have a similarly affected stellar component.

\begin{table*}[t]
\centering
\caption[Derived Properties of Sculptor]{To derive total mass, we use the absolute magnitude from Muñoz et al. (2018) (see also Table \ref{tbl:scl_obs_props}) along with the stellar mass to light ratio (1.7 with 0.17 dex uncertainty) from Woo, Courteau, and Dekel (2008). Sculptor’s total stellar mass is then $M_\star \sim 3.1_{-1.0}^{1.6} \times 10^6\,\Mo$. From the stellar mass to dark matter halo’s characteristic velocity $v_{\rm circ}$ in Fattahi et al. (2018), we expect Sculptor to have $v_{\rm circ} \approx 31 \pm 3 \kms$. }
\label{tbl:scl_derived_props}
\begin{tabular}{lll}
\toprule
parameter & value & reference\\
\midrule
$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & \\
$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
$t_{\rm break, obs}$ & $110\pm30$ Myr & $\sigma_v$, $r_{\rm break}$ with @\\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_initial_halos.pdf}}
\caption[Sculptor initial halos]{The suggested halos of Sculptor
compared to cosmological predictions. \textbf{TODO: simplify
\(\sigma_v\) calculation}.}\label{fig:scl_halos}
\end{figure}

\begin{table*}[t]
\centering
\caption[Initial halos of Sculptor]{The parameters for our initial Sculptor halos. }
\label{tbl:scl_ini_halos}
\begin{tabular}{lllll}
\toprule
Halo name & Rmax & Vmax & M200 & c\\
\midrule
compact & 3.2 & 31 & 0.33 & 21\\
lmc & 4.2 & 31 & 0.39 & 17\\
small & 2.5 & 25 &  & \\
\bottomrule
\end{tabular}
\end{table*}

@@

\subsubsection{Orbital properties}\label{orbital-properties}

Next, we calculate the range of possible orbits of Sculptor in our
fiducial Milky Way potential. Fig.~\ref{fig:scl_orbits} illustrates
point particle orbits for 100 samples of Sculptor's observed kinematics
integrated backwards 5 Gyr in both galactocentric coordinate slices (x,
y, z) and in galactocentric radius with time. In terms of the x, y, z
planes, note that all sampled orbits of Sculptor have nearly the same
morphology---they each orbit mostly in the y-z plane, passing through
the same number of pericentres and maintaining similar orbits across
5Gyr. Sculptor's orbital history is relatively well-constrained.

We select an orbit with the \(\sim 3\sigma\) smallest pericentre among
all possible orbits. We achieve this by taking the median parameters of
all orbits with a pericentre less than the 0.0027th quantile pericentre,
yielding a pericentre of 43 kpc.\footnote{only slightly smaller than the
  0.00135th quantile pericentre} It is unlikely given the current
observations that Sculptor has a smaller pericentre. Since tidal effects
depend most strongly on the pericentre of an orbit, this orbit should
maximize the tidal force from the Milky Way.

We take the first apocentre after a lookback time of 10 Gyr, or at 9.43
Gyr, as the initial conditions for our model of Sculptor.
Table~\ref{tbl:scl_orbits} notes the initial conditions for our
simulation.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_xyzr_orbits.pdf}}
\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
potential in galactocentric \(x\), \(y\), and \(z\) coordinates. The
Milky Way is at the centre with the disk lying in the \(x\)--\(y\)
plane. Our selected \texttt{smallperi} orbit is plotted in black and
light blue transparent orbits represent the past 5Gyr orbits sampled
over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
sculptor is well-constrained in this potential and it is unlikely to
achieve a smaller pericentre than our selected
orbit.}\label{fig:scl_orbits}
\end{figure}

\begin{table*}[t]
\centering
\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor. The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}. The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects. }
\label{tbl:scl_obrits}
\begin{tabular}{llll}
\toprule
Property & Mean & SmallPeri & LMC\\
\midrule
distance & 83.2 & 82.6 & \\
pmra & 0.099 & 0.134 & \\
pmdec & -0.160 & -0.198 & \\
Vlos & 111.2 & 111.2 & \\
$t_i$ & -8.74 & -9.43 & \\
$\hat{x}_{i}$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10] & \\
$\vec{v}_i$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29] & \\
pericentre & 53 & 43 & \\
apocentre & 102 & 96 & \\
$t_{\rm last\ peri}$ & -0.45 & -0.46 & \\
Numer of peris &  &  & \\
\bottomrule
\end{tabular}
\end{table*}

@@

@@

Fig.~\ref{fig:scl_lmc_orbit_effect} displays the effect of including an
LMC in the potential. The green samples are in the initial MW-only
potential in the \texttt{L3M11} model, and the orange samples are
integrated in the evolving MW and LMC \texttt{L3M11} model. The past 1
Gyr is similar in both cases, but the orbits diverge significantly
afterwards. The recent passage of Sculptor with the LMC around 0.1 Gyr
ago allows for Sculptor to begin as far as 300 kpc from the Milky Way
centre. The evolving potential also adds significant long term
variability in the possible orbits of Sculptor. Sculptor, however, is
orbiting in the opposite direction of the LMC so is likely not
associated with the LMC system.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.pdf}}
\caption[Sculptor Orbits with LMC]{This figure is similar to
Fig.~\ref{fig:scl_orbits} except that we are showing the orbits with and
without an LMC. In the bottom row, the distance from Sculptor (or the
LMC) to the MW is plotted (left), and the Sculptor - LMC distance
(right.)}\label{fig:scl_lmc_orbit_effect}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_i_f.pdf}}
\caption[Sculptor initial and final density with LMC]{The tidal effects
on the stellar surface density due to the LMC today.}
\end{figure}

@@

\subsection{Derived properties}\label{derived-properties}

\begin{table*}[t]
\centering
\caption[Derived Properties of Ursa Minor]{Derived properties of Ursa Minor. }
\label{tbl:umi_derived_props}
\begin{tabular}{lll}
\toprule
parameter & value & reference\\
\midrule
$L_\star$ & $3.5 \pm 0.1 \times 10^5\,L_\odot$ & \\
$M_\star$ & $7_{-2}^{+3} \times 10^5\,\Mo$ & \\
$M_\star / L_\star$ & 1.9 (pm 0.17 dex) & Woo, Courteau, and Dekel (2008)\\
$M_{200}$ & $3_{-2}^{+4} \times 10^9\,\Mo$ & \\
$c$ & 14? & \\
$v_{\rm circ, max}$ & $27_{-6}^{+7}\,\kms$ & \\
$r_{\rm circ, max}$ & $5_{-2}^{+1}$ kpc & \\
$r_{\rm break, obs}$ & $30 \pm 5$ arcmin & \\
$t_{\rm break, obs}$ & $120\pm30$ Myr & \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Ursa Minor Initial Halos]{Initial halos for Ursa Minor. }
\label{tbl:umi_ini_halos}
\begin{tabular}{lllll}
\toprule
Halo name & Rmax & Vmax & M200 & c\\
\midrule
fiducial & 5 & 37 & 0.67 & 17\\
compact & 4 & 38 & 0.62 & 21\\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_initial_halos.pdf}}
\caption[Ursa Minor initial halos]{The suggested halos of Ursa Minor
compared to cosmological predictions.}
\end{figure}

\subsection{Orbital properties}\label{orbital-properties-1}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_xyzr_orbits.pdf}}
\caption{Ursa Minor Orbits}
\end{figure}

\subsection{Tidal Effects}\label{tidal-effects-2}

ra = 227.242 dec = 67.2221 distance = 64.6 pmra = -0.158 pmdec = 0.05
radial\_velocity = -245.75

t\_i = -9.53 pericentre = 29.64 apocentre = 74.88 t last peri = -0.80
x\_i = {[}-16.48 69.92 21.05{]} v\_i = {[}16.32 39.86 -116.99{]}

@@

\subsection{Effects of the LMC}\label{effects-of-the-lmc-1}

>>>>

@@

@@

\subsection{Effects of the LMC}\label{sec:scl_lmc}

@@

Fig.~\ref{fig:scl_lmc_orbits_effect} displays the effect of including an
LMC in the potential. The green samples are in the initial MW-only
potential in the \texttt{L3M11} model, and the orange samples are
integrated in the evolving MW and LMC \texttt{L3M11} model. The past 1
Gyr is similar in both cases, but the orbits diverge significantly
afterwards. The recent passage of Sculptor with the LMC around 0.1 Gyr
ago allows for Sculptor to begin as far as 300 kpc from the Milky Way
centre. The evolving potential also adds significant long term
variability in the possible orbits of Sculptor. Sculptor, however, is
orbiting in the opposite direction of the LMC so is likely not
associated with the LMC system.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.pdf}}
\caption[Sculptor Orbits with LMC]{This figure is similar to
Fig.~\ref{fig:scl_orbits} except that we are showing the orbits with and
without an LMC. In the bottom row, the distance from Sculptor (or the
LMC) to the MW is plotted (left), and the Sculptor - LMC distance
(right.)}\label{fig:scl_lmc_orbits_effect}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_i_f.pdf}}
\caption[Sculptor initial and final density with LMC]{The tidal effects
on the stellar surface density due to the LMC
today.}\label{fig:scl_lmc_i_f}
\end{figure}

@@

@@
